---
title: "Adult Depression in California through 2012 to 2018"
author: Ashas Memon
thanks: "Code and data are available at: https://github.com/ashasmemon/MentalHealthExploration"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "This report aims to examine Adult Depression in California as an indicator in Lets’ Get Healthy California including a general breakdown of age, sex, ethnicity and education. It discusses data collection methodology and examines the changes in depression rates among adults from various backgrounds ands education levels from 2012 to 2018. The data examines here was collected by asking, “Has a doctor, nurse or other health care professional ever told you that you have depressive disorder?”"
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE, show_col_types = FALSE}

# install.packages("haven")
# install.packages("tidyverse")
# install.packages("dplyr")
# install.packages("janitor")
# install.packages("tidyr")
# install.packages("ggplot2")
# install.packages("lubridate")
# install.packages("gridExtra")
# install.packages("scales")
# install.packages("knitr")
library(haven)
library(tidyverse)
library(dplyr)
library(janitor) # Helps clean datasets
library(tidyr) # Helps make tidy datasets
library(ggplot2)
library(lubridate)
library(gridExtra)
library(scales)
library(knitr)
library(kableExtra)

knitr::opts_chunk$set(echo = TRUE)

raw_data <- readr::read_csv("../../inputs/data/indicators.csv")
sex_col <- select(filter(raw_data, Strata == "Sex"), c(1:4))
names(sex_col)[3] <- "sex"

education_col <- select(filter(raw_data, Strata == "Education"), c(1:4))


age_col <- select(filter(raw_data, Strata == "Age"), c(1:4))
names(age_col)[3] <- "age"

race_col <- select(filter(raw_data, Strata == "Race-Ethnicity"), c(1:4))
names(race_col)[3] <- "race"
```

# Introduction

Over the past 3 years with the economic recession due to the Covid-19 pandemic, there has been an increase in the number of people affected by some type of mental health disorder. As the pandemic progressed, there has been a significant increases in reported cases of individuals and households with anxiety or depressive disorder mostly attributed to the prolonged public health measures (https://www.kff.org/coronavirus-covid-19/issue-brief/the-implications-of-covid-19-for-mental-health-and-substance-use/ ).  The data was collected as an indicator in Lets’ Get Healthy California and includes a general breakdown of age, sex, ethnicity, education (among other strata) of individuals who reported they had a depressive disorder. The data was collected as part of an annual health related telephone survey conducted for California Behavioral Risk Factor Surveillance Survey (BRFSS), where individuals were asked, “Has a doctor, nurse or other health care professional EVER told you that you have a depressive disorder (including depression, major depression, dysthymia or minor depression)?”. The frequencies of all these individuals who responded to this was recorded in multiple tables. It is important to note that this particular dataset is not affiliated with the Centre for Disease Control and Prevention (CDC) or any other federal agencies, it is a stand-alone set and only contains data from 2012 up to 2018.

# Data


## Data Summary



## R Packages

This paper uses R[@citeR] for the main analysis of this paper and various other packages including **tidyverse** [@tidyverse], **dplyr** [@dplyr], **janitor** [@janitor], and **tidyr** [@tidyr] for cleaning, processing and manipulating the data. Additionally, **ggplot2** [@ggplot], and **gridExtra** [@gridExtra] packages are used to to create appropriate plots and visual representations of the data presented throughout this report.


# Methodology

## Sampling Population 

The data was collected as an indicator in Lets’ Get Healthy California and includes a general breakdown of age, sex, ethnicity, education (among other strata) of individuals who reported they had a depressive disorder. The sampling population is adults living in California USA above the age of 18.

## Sampling Method

The data was collected as part of an annual health related telephone survey conducted for California Behavioral Risk Factor Surveillance Survey (BRFSS), where individuals were asked, “Has a doctor, nurse or other health care professional EVER told you that you have a depressive disorder (including depression, major depression, dysthymia or minor depression)?”. The frequencies of all these individuals who responded to this was recorded in multiple tables. It is important to note that this particular dataset is not affiliated with the Centre for Disease Control and Prevention (CDC) or any other federal agencies, it is a stand-alone set and only contains data from 2012 up to 2018.

# Findings


```{r sexLinePlot, fig.cap="Distribution from Males vs Females", echo=FALSE, warning=FALSE,show_col_types = FALSE}

sex_line_plot <- ggplot(sex_col, aes(x = Year, y = Frequency, fill = sex)) +
  geom_line() +
  geom_point(size = 4, shape = 21) +
  labs(x="Year", y="Number of Cases", title="Distribution of Depression for Male vs Females")

sex_line_plot
```

From the figure \@ref(fig:sexLinePlot), the number of cases amoung males increases slightly overall where as the number of cases in women decreases sligtly over the course from 2012 to 2018. The line representing both males and females has a dip in 2014 with less cases reported during that year and a sudden increase in the following year and then another drop in the number of cases in 2016. For males, the number of cases decrease from around 600 to just above 400 from 2012 to 2014 and the number of cases in women starts at just under 1400 followed by a significant decrease to around 800 in 2014. 

It is also clear from figure \@ref(fig:sexLinePlot) that there are significantly higher depression cases among females then there are amoung males. There is at least a 300 case difference in the number of cases in females as compared to males. 

\newpage

```{r raceBarPlot, fig.cap="Distribution pf Depression by Race", echo=FALSE, warning=FALSE,show_col_types = FALSE}

race_bar_plot <- ggplot(race_col, aes(x = Year, y = Frequency, fill = race)) +
  geom_col(position = position_dodge()) +
  labs(x="Year", y="Number of Cases", title="Distribution of Depression by Race")

race_bar_plot
```


After plotting the number of cases of depressive disorder by race and ethnic background in figure \@ref(fig:raceBarPlot), there is a significantly higher number of depression cases amoung White individuals than any other race. For each of the years between 2012 to 2018, number of cases amoung White individuals seems to decrease slightly on average, although it is still higher than all other racial backgrounds for each of the given years. Additionally, Hispanics have the second largest number of cases in each year but significantly less than White and significantly higher than the other racial backgrounds.

Upon further examination, Asian/Pacific Islanders have the lowest number of depression cases with under 100 cases for each of the years. The second highest number of cases is amound the Black community with number of cases just slightly higher than Asian/Pacific Islanders. There is a drop in the number of cases in 2014 for all races, followed by an increase in 2015, which is again followed by a decrease in 2016.

\newpage

```{r ageLinePlot, fig.cap="Distribution of Depression by Age", echo=FALSE, warning=FALSE,show_col_types = FALSE}

age_bar_plot <- ggplot(age_col, aes(x = Year, y = Frequency, fill = age)) +
  geom_col(position = position_dodge()) +
  labs(x="Year", y="Number of Cases", title="Distribution of Depression by Age")

age_line_plot <- ggplot(age_col, aes(x = Year, y = Frequency, color = age)) +
  geom_line() +
  geom_point(size = 2, shape = 21) +
  labs(x="Year", y="Number of Cases", title="Distribution of Depression by Age")

age_line_plot

# age_bar_plot
```

 
As seen in figure \@ref(fig:ageLinePlot), there is a different trend for each age group. There is an overall increase in the number of cases from 2012 to 2018 for age group 18 to 34. There is also a slight overall increase in the number of cases from 2012 to 2018 amoung the age group 35 to 44 but not a significantly increasing trend as seen in age group 18 to 34. It is also important to note that most of the age intervals cover 10 year intervals except for age group 18 to 34 and above 65. Additionally, the data might be slightly different because figure \@ref(fig:ageLinePlot) represents frequencies instead of normalized results for each of the age groups. This could account for significantly higher number amoung 18 to 34 age group because there are more individuals in that age range. Furthermore, the overall number of cases in the age groups 45 to 54, 55 to 64 and 65+ appears to decrease slighly over the years 2012 to 2018. These three groups appear to show the same trend but the only different being that the age group of 45 to 54 having around 100 cases less for each year and ther other two groups having around the same number of cases for each year.

There is a drop in the number of reported cases in 2014 and the following year (2015) has a sudden increase amoung all the age groups, followed again by a drop in the number of cases in 2016. This trend of dropping number of cases in 2014, increase in 2015 and dropping again in 2016 is seen in all of the data collected in figure \@ref(fig:sexLinePlot) and \@ref(fig:raceBarPlot) as well.
 
# Discussion Based on Findings 

## Sex

Upon further examining of literature about the gender gap is depression, women are twice as likely to depression as compared to men [@GenderGapOne]. Higher depression cases in females are caused by a number of other factors as well, including hormonal changes, inherited traits and lifestyle. Sometimes hormonal changes may cause short periods of temporary depression but most likely nothing long term. Puberty is one cause where there are a lot of changes going on in ones life including sexual and identity issues, and pressure from family to achieve in school and other aspects of life [@GenderGapOne]. 

Studies have also suggested that women may experience more stressfull life events and they are more sensitive to them as compared to men. Furthermore, its is also seen that depression rates are fairly consistent throughout the world (when comparing men to women) which could suggest that underlying biological factors play a major role in determining in addition to other factors such as social status, life experience, race, etc [@GenderGapTwo].

## Race

Hispanic and black communities still view depression as a taboo topic and are less likely to report those cases if the are asked. Culturally, they don't believe that depression is a disorder so they don't admit it. There are also a number of other reasons why Hispanics don't seek help for those who show symptoms of depression, including language barriers, lack of family involvement and cultural influences [@HispanicCitation]. People who dont speak the language of their region very clearly are not as likely to seek help because of the language barrier. Additionally, there is still a large family and cultural impact in the Hispanic community that view mental health disorders as a stigma topic so a lot of of people with depression don't speak out in fear of their family not acknowledging their situation [@HispanicCitation]. According to [@CaliforniaRaceCitation], there are almost an equal number of White and Hispanic individuals living in California with their percentages being $38.4$% and $38.1$% respectively. With such a close number of these two racial groups, we would expect there to be around the same number of depression cases but that is not the case, we see significantly higher reported cases amount white individuals compared to Hispanics.

In addition to that, around $5.9$ % of California's population is Black and $13.9$ % is Asian but in the results presented in figure \@ref(fig:raceBarPlot), there are more reported cases amoung Blacks than Asians even though there are almost double the number of people of Asian background. 

## Age



# Privacy and Ethics 

Data was collected anonymously about racial background and other factors including age,sex but not any personal information to link the data back to any of the surveyed individuals.

\newpage

# Appendix

## Datasheet 
Datasheet was obtained from link below and modified slightly for the use in this report.

Extract of the questions from [@gebru2021datasheets]

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The data was collected from Data.world about the number of cases of Adult depression in California.
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - The data was collected by California Behavioral Risk Factor Surveillance Survey (BRFSS) in California but it is not a part of the Center for Disease and Control (CDC).

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- Not applicable
2. *How many instances are there in total (of each type, if appropriate)?*
	- Just 1
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- No, this is a sample phone survey, individuals were selected randomly 
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- Data contains frequencies of individuals diagnosed with depression disorder along with that individuals age, racial background, income, education and other such factors across the years 2012-2018.
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- Not Applicable
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- No.
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- No
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- No
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- Not that are immediately appearant
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- Self-contained
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- No, data was collected anonymously
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- No
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- sub populations are divided by age, race, education level, income to name a few.
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- No, data is anonymously collected.
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- No
16. *Any other comments?*
	- No.

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- phone survey
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- Phone survey
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- SUrvey data, so a sample
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- California Behavioral Risk Factor Surveillance Survey (BRFSS) and individuals associated to that.
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- collected from 2012 to 2018
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- None that is appearent.
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- from data.world website
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- data collected by phone survey
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- Yes
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- Not to my knowledge.
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- Not that i am aware of.
12. *Any other comments?*
	- No.

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- already well labled
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- No
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- RStudio
4. *Any other comments?*
	- No

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- a few online open source studies founf on data.world under this data set about Adult Depression in California
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- https://data.world/chhs/5a281abf-1730-43b0-b17b-ac6a35db5760 
3. *What (other) tasks could the dataset be used for?*
	- Other explorations related to adult depression 
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- No
5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- No
6. *Any other comments?*
	- No

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- TBD
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- TBD
3. *When will the dataset be distributed?*
	- TBD
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- TBD
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- TBD
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- TBD
7. *Any other comments?*
	- TBD

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- https://data.world/chhs/5a281abf-1730-43b0-b17b-ac6a35db5760
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- Contact Lets Get Healthy California: https://letsgethealthy.ca.gov/.
3. *Is there an erratum? If so, please provide a link or other access point.*
	- No.
4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- Not since 2019
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- Not sure.
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- Data was posted in 2019 and has not been updated since.
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- Contact the survey organization directly and finalize with them.
8. *Any other comments?*
	- No.


\newpage

\newpage

# Reference
