---
title: "Adult Depression in California through 2012 to 2018"
author: Ashas Memon
thanks: "Code and data are available at: https://github.com/ashasmemon/MentalHealthExploration"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "This report aims to examine Adult Depression in California as an indicator in Lets’ Get Healthy California including a general breakdown of age, sex, ethnicity and education. It discusses data collection methodology and examines the changes in depression rates among adults from various backgrounds ands education levels from 2012 to 2018. The data examines here was collected by asking, “Has a doctor, nurse or other health care professional ever told you that you have depressive disorder?”"
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE, show_col_types = FALSE}

# install.packages("haven")
# install.packages("tidyverse")
# install.packages("dplyr")
# install.packages("janitor")
# install.packages("tidyr")
# install.packages("ggplot2")
# install.packages("lubridate")
# install.packages("gridExtra")
# install.packages("scales")
# install.packages("knitr")
library(haven)
library(tidyverse)
library(dplyr)
library(janitor) # Helps clean datasets
library(tidyr) # Helps make tidy datasets
library(ggplot2)
library(lubridate)
library(gridExtra)
library(scales)
library(knitr)
library(kableExtra)

knitr::opts_chunk$set(echo = TRUE)

raw_data <- readr::read_csv("../../inputs/data/indicators.csv")
sex_col <- select(filter(raw_data, Strata == "Sex"), c(1:4))
names(sex_col)[3] <- "sex"

education_col <- select(filter(raw_data, Strata == "Education"), c(1:4))


age_col <- select(filter(raw_data, Strata == "Age"), c(1:4))
names(age_col)[3] <- "age"

race_col <- select(filter(raw_data, Strata == "Race-Ethnicity"), c(1:4))
names(race_col)[3] <- "race"
```

# Introduction

Over the past 3 years with the economic recession due to the Covid-19 pandemic, there has been an increase in the number of people affected by some type of mental health disorder. As the pandemic progressed, there has been a significant increases in reported cases of individuals and households with anxiety or depressive disorder mostly attributed to the prolonged public health measures (https://www.kff.org/coronavirus-covid-19/issue-brief/the-implications-of-covid-19-for-mental-health-and-substance-use/ ).  The data was collected as an indicator in Lets’ Get Healthy California and includes a general breakdown of age, sex, ethnicity, education (among other strata) of individuals who reported they had a depressive disorder. The data was collected as part of an annual health related telephone survey conducted for California Behavioral Risk Factor Surveillance Survey (BRFSS), where individuals were asked, “Has a doctor, nurse or other health care professional EVER told you that you have a depressive disorder (including depression, major depression, dysthymia or minor depression)?”. The frequencies of all these individuals who responded to this was recorded in multiple tables. It is important to note that this particular dataset is not affiliated with the Centre for Disease Control and Prevention (CDC) or any other federal agencies, it is a stand-alone set and only contains data from 2012 up to 2018.

# Data


## Data Summary



## R Packages

This paper uses R[@citeR] for the main analysis of this paper and various other packages including **tidyverse** [@tidyverse], **dplyr** [@dplyr], **janitor** [@janitor], and **tidyr** [@tidyr] for cleaning, processing and manipulating the data. Additionally, **ggplot2** [@ggplot], and **gridExtra** [@gridExtra] packages are used to to create appropriate plots and visual representations of the data presented throughout this report.


# Methodology

## Features

## Sampling Population 

## Sampling Method

# Findings



```{r sexLinePlot, fig.cap="Distribution from Males vs Females", echo=FALSE, warning=FALSE,show_col_types = FALSE}

sex_line_plot <- ggplot(sex_col, aes(x = Year, y = Frequency, fill = sex)) +
  geom_line() +
  geom_point(size = 4, shape = 21) +
  labs(x="Year", y="Number of Cases", title="Distribution of Depression for Male vs Females")

sex_line_plot
```

From the figure \@ref(fig:sexLinePlot), the number of cases amoung males increases slightly overall where as the number of cases in women decreases sligtly over the course from 2012 to 2018. The line repreentaing both males and females has a dip in 2014 with less cases reported during that year and a sudden increase in the following year. For make, the number of cases decrease from around 600 to just above 400 from 2012 to 2014 and the number of cases in women starts at just under 1400 followed by a significant decrease to around 800 in 2014.


```{r raceBarPlot, fig.cap="Distribution pf Depression by Race", echo=FALSE, warning=FALSE,show_col_types = FALSE}

race_bar_plot <- ggplot(race_col, aes(x = Year, y = Frequency, fill = race)) +
  geom_col(position = position_dodge()) +
  labs(x="Year", y="Number of Cases", title="Distribution of Depression by Race")

race_bar_plot
```


As seen in figure \@ref(fig:raceBarPlot), there is a significantly higher number of depression amoung white individuals that any other race.

```{r ageLinePlot, fig.cap="Distribution of Depression by Age", echo=FALSE, warning=FALSE,show_col_types = FALSE}

age_bar_plot <- ggplot(age_col, aes(x = Year, y = Frequency, fill = age)) +
  geom_col(position = position_dodge()) +
  labs(x="Year", y="Number of Cases", title="Distribution of Depression by Age")

age_line_plot <- ggplot(age_col, aes(x = Year, y = Frequency, color = age)) +
  geom_line() +
  geom_point(size = 2, shape = 21) +
  labs(x="Year", y="Number of Cases", title="Distribution of Depression by Age")

age_line_plot

# age_bar_plot
```

 
 As seen in figure \@ref(fig:ageLinePlot), there is a a differnt trend for each age group.
 
# Discussion Based on Findings 

# Privacy and Ethics 

# Appendix

\newpage

# Reference
