---
title: "Adult Depression in California through 2012 to 2018"
author: Ashas Memon
thanks: "Code and data are available at: https://github.com/ashasmemon/MentalHealthExploration"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "This report aims to examine Adult Depression in California as an indicator in Lets’ Get Healthy California including a general breakdown of age, sex, ethnicity and education. It discusses data collection methodology and examines the changes in depression rates among adults from various backgrounds ands education levels from 2012 to 2018. The data examines here was collected by asking, “Has a doctor, nurse or other health care professional ever told you that you have depressive disorder?”"
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE, show_col_types = FALSE}

# install.packages("haven")
# install.packages("tidyverse")
# install.packages("dplyr")
# install.packages("janitor")
# install.packages("tidyr")
# install.packages("ggplot2")
# install.packages("lubridate")
# install.packages("gridExtra")
# install.packages("scales")
# install.packages("knitr")
library(haven)
library(tidyverse)
library(dplyr)
library(janitor) # Helps clean datasets
library(tidyr) # Helps make tidy datasets
library(ggplot2)
library(lubridate)
library(gridExtra)
library(scales)
library(knitr)
library(kableExtra)

knitr::opts_chunk$set(echo = TRUE)

raw_data <- readr::read_csv("../../inputs/data/indicators.csv")
sex_col <- select(filter(raw_data, Strata == "Sex"), c(1:4))
names(sex_col)[3] <- "sex"

education_col <- select(filter(raw_data, Strata == "Education"), c(1:4))


age_col <- select(filter(raw_data, Strata == "Age"), c(1:4))
names(age_col)[3] <- "age"

race_col <- select(filter(raw_data, Strata == "Race-Ethnicity"), c(1:4))
names(race_col)[3] <- "race"
```

# Introduction

Over the past 3 years with the economic recession due to the Covid-19 pandemic, there has been an increase in the number of people affected by some type of mental health disorder. As the pandemic progressed, there has been a significant increases in reported cases of individuals and households with anxiety or depressive disorder mostly attributed to the prolonged public health measures (https://www.kff.org/coronavirus-covid-19/issue-brief/the-implications-of-covid-19-for-mental-health-and-substance-use/ ).  The data was collected as an indicator in Lets’ Get Healthy California and includes a general breakdown of age, sex, ethnicity, education (among other strata) of individuals who reported they had a depressive disorder. The data was collected as part of an annual health related telephone survey conducted for California Behavioral Risk Factor Surveillance Survey (BRFSS), where individuals were asked, “Has a doctor, nurse or other health care professional EVER told you that you have a depressive disorder (including depression, major depression, dysthymia or minor depression)?”. The frequencies of all these individuals who responded to this was recorded in multiple tables. It is important to note that this particular dataset is not affiliated with the Centre for Disease Control and Prevention (CDC) or any other federal agencies, it is a stand-alone set and only contains data from 2012 up to 2018.

# Data


## Data Summary



## R Packages

This paper uses R[@citeR] for the main analysis of this paper and various other packages including **tidyverse** [@tidyverse], **dplyr** [@dplyr], **janitor** [@janitor], and **tidyr** [@tidyr] for cleaning, processing and manipulating the data. Additionally, **ggplot2** [@ggplot], and **gridExtra** [@gridExtra] packages are used to to create appropriate plots and visual representations of the data presented throughout this report.


# Methodology

## Features

## Sampling Population 

## Sampling Method

# Findings



```{r sexLinePlot, fig.cap="Distribution from Males vs Females", echo=FALSE, warning=FALSE,show_col_types = FALSE}

sex_line_plot <- ggplot(sex_col, aes(x = Year, y = Frequency, fill = sex)) +
  geom_line() +
  geom_point(size = 4, shape = 21) +
  labs(x="Year", y="Number of Cases", title="Distribution of Depression for Male vs Females")

sex_line_plot
```

From the figure \@ref(fig:sexLinePlot), the number of cases amoung males increases slightly overall where as the number of cases in women decreases sligtly over the course from 2012 to 2018. The line representing both males and females has a dip in 2014 with less cases reported during that year and a sudden increase in the following year and then another drop in the number of cases in 2016. For males, the number of cases decrease from around 600 to just above 400 from 2012 to 2014 and the number of cases in women starts at just under 1400 followed by a significant decrease to around 800 in 2014. 

It is also clear from figure \@ref(fig:sexLinePlot) that there are significantly higher depression cases among females then there are amoung males. There is at least a 300 case difference in the number of cases in females as compared to males.

\newpage

```{r raceBarPlot, fig.cap="Distribution pf Depression by Race", echo=FALSE, warning=FALSE,show_col_types = FALSE}

race_bar_plot <- ggplot(race_col, aes(x = Year, y = Frequency, fill = race)) +
  geom_col(position = position_dodge()) +
  labs(x="Year", y="Number of Cases", title="Distribution of Depression by Race")

race_bar_plot
```


After plotting the number of cases of depressive disorder by race and ethnic background in figure \@ref(fig:raceBarPlot), there is a significantly higher number of depression cases amoung White individuals than any other race. For each of the years between 2012 to 2018, number of cases amoung White individuals seems to decrease slightly on average, although it is still higher than all other racial backgrounds for each of the given years. Additionally, Hispanics have the second largest number of cases in each year but significantly less than White and significantly higher than the other racial backgrounds.

Upon further examination, Asian/Pacific Islanders have the lowest number of depression cases with under 100 cases for each of the years. The second highest number of cases is amound the Black community with number of cases just slightly higher than Asian/Pacific Islanders. There is a drop in the number of cases in 2014 for all races, followed by an increase in 2015, which is again followed by a decrease in 2016.

\newpage

```{r ageLinePlot, fig.cap="Distribution of Depression by Age", echo=FALSE, warning=FALSE,show_col_types = FALSE}

age_bar_plot <- ggplot(age_col, aes(x = Year, y = Frequency, fill = age)) +
  geom_col(position = position_dodge()) +
  labs(x="Year", y="Number of Cases", title="Distribution of Depression by Age")

age_line_plot <- ggplot(age_col, aes(x = Year, y = Frequency, color = age)) +
  geom_line() +
  geom_point(size = 2, shape = 21) +
  labs(x="Year", y="Number of Cases", title="Distribution of Depression by Age")

age_line_plot

# age_bar_plot
```

 
As seen in figure \@ref(fig:ageLinePlot), there is a different trend for each age group. There is an overall increase in the number of cases from 2012 to 2018 for age group 18 to 34. There is also a slight overall increase in the number of cases from 2012 to 2018 amoung the age group 35 to 44 but not a significantly increasing trend as seen in age group 18 to 34. It is also important to note that most of the age intervals cover 10 year intervals except for age group 18 to 34 and above 65. Additionally, the data might be slightly different because figure \@ref(fig:ageLinePlot) represents frequencies instead of normalized results for each of the age groups. This could account for significantly higher number amoung 18 to 34 age group because there are more individuals in that age range. Furthermore, the overall number of cases in the age groups 45 to 54, 55 to 64 and 65+ appears to decrease slighly over the years 2012 to 2018. These three groups appear to show the same trend but the only different being that the age group of 45 to 54 having around 100 cases less for each year and ther other two groups having around the same number of cases for each year.

There is a drop in the number of reported cases in 2014 and the following year (2015) has a sudden increase amoung all the age groups, followed again by a drop in the number of cases in 2016. This trend of dropping number of cases in 2014, increase in 2015 and dropping again in 2016 is seen in all of the data collected in figure \@ref(fig:sexLinePlot) and \@ref(fig:raceBarPlot) as well.
 
# Discussion Based on Findings 

## Sex

Upon further examining of literature about the gender gap is depression, women are twice as likely to depression as compared to men [@GenderGapOne]. Higher depression cases in females are caused by a number of other factors as well, including hormonal changes, inherited traits and lifestyle. Sometimes hormonal changes may cause short periods of temporary depression but most likely nothing long term. Puberty is one cause where there are a lot of changes going on in ones life including sexual and identity issues, and pressure from family to achieve in school and other aspects of life [@GenderGapOne]. 

Studies have also suggested that women may experience more stressfull life events and they are more sensitive to them as compared to men. Furthermore, its is also seen that depression rates are fairly consistent throughout the world (when comparing men to women) which could suggest that underlying biological factors play a major role in determining in addition to other factors such as social status, life experience, race, etc [@GenderGapTwo].

## Race

Hispanic and black communities still view depression as a taboo topic and are less likely to report those cases if the are asked. Culturally, they don't believe that depression is a disorder so they don't admit it. could also be that there are significantly less Hispanic and black individuals living in California so that why there are less cases.

## Age

# Privacy and Ethics 

Data collected anonymously, data collection about racial background and other factors including age,sex but not any personal information to link the data back to any of the surveyed individuals.

# Appendix

\newpage

# Reference
